---
title: "Difference in Dialysis Patient Care Between Large Chain Clinics and Independent Firms"
author: "Rahul Narvekar, Alex Ozhakanat, Alison Zhang, Shivangi Shah, Justin Geletko"
date: "12/09/2021"
output:
  pdf_document: default
  html_document: default
---

# Downloading Data
Source: National Bureau of Economic Research
```{r download-data}
# check to see if the first year exists
exists = file.exists("hcris_raw/rpt1998.csv")
if (!exists) {
  for(i in 1998:2010) {
    # download into local folder
    download.file(paste0("http://www.nber.org/hcris/265-94/rnl_rpt265_94_",i,".csv"), paste0("hcris_raw/rpt",i,".csv"))
    download.file(paste0("http://www.nber.org/hcris/265-94/rnl_nmrc265_94_",i,"_long.csv"), paste0("hcris_raw/nmrc",i,".csv"))
    download.file(paste0("http://www.nber.org/hcris/265-94/rnl_alpha265_94_",i,"_long.csv"), paste0("hcris_raw/alpha",i,".csv"))
  }
}
print("data downloaded!")
```


# Merging Data
```{r data-generation}

# helper function made the name variable features
colNameMaker = function(original) {
  return(strsplit(original, split = "[.]")[[1]][2])
}

# if the file for the csv does not exist, go ahead and generate it
if(!file.exists("hcris_raw/HcrisPanel.csv")) {
  library(stringr)
  
  # name of all features
  allFeatures = c("rpt_rec_num","facility_name","non_medicare_sessions",         
                  "non_medicare_sessions_indirect","avg_weekly_sessions","avg_days_open_per_week",      
                  "avg_session_time","num_machines_regular","num_machines_standby",          
                  "dialyzer_type","dialyser_reuse_times","epo_total",                     
                  "epo_cost" ,"epo_rebates","epo_net_cost",                  
                  "chain_indicator","chain_identity","prvdr_num",                     
                  "ever_hospital_based","certification_date","report_start_date",            
                  "report_end_date","total_treatments_hd","total_costs_hd_housekeeping",   
                  "total_costs_hd_machines","total_costs_hd_salaries","total_costs_hd_benefits",       
                  "total_costs_hd_drugs","total_costs_hd_supplies","total_costs_hd_labs",           
                  "total_costs_hd_other","supplies","lab_services",                  
                  "total_treatments_pd","zip_code","state",                         
                  "fy_bgn_dt","fy_end_dt","year")
  
  # initialize a blank data frame
  HcrisPanel =  data.frame(matrix(ncol=39, nrow = 0))
  colnames(HcrisPanel) = allFeatures
  
  # generate keys from var code csv
  varCodes = read.csv('hcris_raw/variable_codes.csv')
  keysVarCodes = rep("", length(varCodes))
  for(i in 1:NROW(varCodes)) {
    worksheet = varCodes[i,2]
    line = str_pad(varCodes[i,3], 5, pad = "0")
    column = str_pad(varCodes[i,4], 4, pad = "0")
    keysVarCodes[i] = paste0(worksheet, line, column)
  }
  varCodes$keys = keysVarCodes
  
  # loop through every year and create a data frame to rbind to HcrisPanel
  for (j  in 1998:2010) {
    
    # read in alpha, nmrc, and rpt files for current year
    alpha = read.csv(paste0('hcris_raw/alpha',j,'.csv'))
    nmrc = read.csv(paste0('hcris_raw/nmrc',j,'.csv'))
    rpt = read.csv(paste0('hcris_raw/rpt',j,'.csv'))
    keysAlpha = rep("", length(alpha))
    keysNmrc = rep("", length(nmrc))
    
    # generate keys for nmrc and alpha
    for(i in 1:NROW(nmrc)) {
      worksheet = nmrc[i,2]
      line = str_pad(nmrc[i,3], 5, pad = "0")
      column = str_pad(nmrc[i,4], 4, pad = "0")
      keysNmrc[i] = paste0(worksheet, line, column)
    }
    nmrc$keys = keysNmrc
    for(i in 1:NROW(alpha)) {
      worksheet = alpha[i,2]
      line = str_pad(alpha[i,3], 5, pad = "0")
      column = str_pad(alpha[i,4], 4, pad = "0")
      keysAlpha[i] = paste0(worksheet, line, column)
    }
    alpha$keys = keysAlpha
    
    # merge var nmrc and alpha with var code
    alphaWithVars = merge(varCodes[,c(1,5)], alpha[,c(1,5,6)], by = 'keys', all.x = TRUE)
    nmrcWithVars = merge(varCodes[,c(1,5)], nmrc[,c(1,5,6)], by = 'keys', all.x = TRUE)
    
    #reshape data frames from long to wide
    wideAlpha = reshape(alphaWithVars[,-1], idvar = "rpt_rec_num", timevar = "variable", direction = "wide")
    wideNmrc = reshape(nmrcWithVars[,-1], idvar = "rpt_rec_num", timevar = "variable", direction = "wide")
    screwedUpMerge = merge(wideAlpha, wideNmrc, by = 'rpt_rec_num', all.x = T)
    
    # grab required features
    features = c("rpt_rec_num", "alphnmrc_itm_txt.facility_name","itm_val_num.non_medicare_sessions",
      "itm_val_num.non_medicare_sessions_indirect","itm_val_num.avg_weekly_sessions","itm_val_num.avg_days_open_per_week",
      "itm_val_num.avg_session_time","itm_val_num.num_machines_regular","itm_val_num.num_machines_standby",        
      "itm_val_num.dialyzer_type","itm_val_num.dialyser_reuse_times","itm_val_num.epo_total",                
      "itm_val_num.epo_cost","itm_val_num.epo_rebates","itm_val_num.epo_net_cost",            
      "alphnmrc_itm_txt.chain_indicator","alphnmrc_itm_txt.chain_identity","alphnmrc_itm_txt.prvdr_num",            
      "alphnmrc_itm_txt.ever_hospital_based","alphnmrc_itm_txt.certification_date","alphnmrc_itm_txt.report_start_date",     
      "alphnmrc_itm_txt.report_end_date","itm_val_num.total_treatments_hd","itm_val_num.total_costs_hd_housekeeping",
      "itm_val_num.total_costs_hd_machines","itm_val_num.total_costs_hd_salaries","itm_val_num.total_costs_hd_benefits", 
      "itm_val_num.total_costs_hd_drugs","itm_val_num.total_costs_hd_supplies","itm_val_num.total_costs_hd_labs",   
      "itm_val_num.total_costs_hd_other","itm_val_num.supplies","itm_val_num.lab_services",           
      "itm_val_num.total_treatments_pd","alphnmrc_itm_txt.zip_code","alphnmrc_itm_txt.state")
    cleanedMerge = screwedUpMerge[features]
    
    # drop item_val_nul and alphnmrc_itm_text from variable names
    for(i in 2:length(features)) {
      features[i] = colNameMaker(features[i])
    }
    colnames(cleanedMerge) = features
    
    # merge in fy bgn and end date from rpt data set
    currentYear = merge(cleanedMerge, rpt[,c(1,13,14)], by = 'rpt_rec_num', all.y = TRUE)
    
    # add current year
    currentYear$year = j
    
    # rbind HcrisPanel with the current year
    HcrisPanel = rbind(HcrisPanel, currentYear)
  }
  
  # write out the final csv file
  write.csv(HcrisPanel, "hcris_raw/HcrisPanel.csv", row.names = FALSE)
}

# read in merged csv file
HcrisPanel = read.csv("hcris_raw/HcrisPanel.csv")
print("data frames merged!")
```

# Cleaning Data
```{r data-cleaning}
library(stringr)
library(tidyverse)
library(lubridate)

raw_data = HcrisPanel

# drop NA prvdr_num rows
d = raw_data %>% drop_na(prvdr_num)

# account for negative epo variable records
d$epo_cost = abs(d$epo_cost)
d$epo_net_cost = abs(d$epo_net_cost)
d$epo_rebates = abs(d$epo_rebates)

# replace missing values with 0 for epo_rebates
d$epo_rebates = replace(d$epo_rebates, is.na(d$epo_rebates), 0)

# check missing epo_cost and epo_net_cost observations
d$epo_cost = ifelse((is.na(d$epo_cost) & !is.na(d$epo_net_cost) & d$epo_rebates == 0), d$epo_net_cost, d$epo_cost)

d$epo_cost = ifelse((is.na(d$epo_cost) & !is.na(d$epo_net_cost) & d$epo_rebates != 0), d$epo_net_cost + d$epo_rebates, d$epo_cost)

blank_idxes = rep(NA, 1)

for (x in 1:length(d$epo_cost)) {
  if (is.na(d$epo_cost[[x]]) & is.na(d$epo_net_cost[[x]])) {
    if (d$epo_rebates[[x]] == 0) {
      d$epo_cost[[x]] = 0
      d$epo_net_cost[[x]] = 0
    } else {
      blank_idxes = c(blank_idxes, x)
    }
  }
}

d$epo_net_cost = ifelse(!is.na(d$epo_cost) & is.na(d$epo_net_cost), d$epo_cost - d$epo_rebates, d$epo_net_cost)

# switch epo_cost and epo_net_cost
for (x in 1:length(d$epo_cost)) {
  if (!is.na(d$epo_cost[[x]]) & !is.na(d$epo_net_cost[[x]]) & d$epo_cost[[x]] < d$epo_net_cost[[x]]) {
    tmp = d$epo_cost[[x]]
    d$epo_cost[[x]] = d$epo_net_cost[[x]]
    d$epo_net_cost[[x]] = tmp
  }
}

# change prvdr_num from “322664” to "342664"
d$prvdr_num = ifelse(d$prvdr_num == 322664, 342664, d$prvdr_num)

# convert fy_bgn_dt, fy_end_dt, report_start_date, and report_end_date to date format
d$fy_bgn_dt = mdy(d$fy_bgn_dt)
d$fy_end_dt = mdy(d$fy_end_dt)
d$report_start_date = mdy(d$report_start_date)
d$report_end_date = mdy(d$report_end_date)

# use fy_bgn_dt and fy_end_dt for missing report start and end dates
for (x in 1:length(d$report_start_date)) {
  if (is.na(d$report_start_date[[x]])) {
    d$report_start_date[[x]] = d$fy_bgn_dt[[x]];
  }
  
  if (is.na(d$report_end_date[[x]])) {
    d$report_end_date[[x]] = d$fy_end_dt[[x]];
  }
}

# reformat zip codes
for (x in 1:length(d$zip_code)) {
trimmed = trimws(d$zip_code[[x]], whitespace = "[\t\r\n]")
d$zip_code[[x]] = as.numeric(substr(trimmed, 1, 5))
}

# use mode to replace missing zip codes for each provider number
getMode = function(v) {
uniqv = unique(v)
uniqv[which.max(tabulate(match(v, uniqv)))]
}

providers = unique(d$prvdr_num, incomparables = c(NA))

for (p in providers) {
prvdr_zips = subset(d, d$prvdr_num == p)$zip_code
prvdr_zips = prvdr_zips[!is.na(prvdr_zips)]

if (length(prvdr_zips) > 0) {
curr_zip = getMode(prvdr_zips)
print(curr_zip)

d$zip_code = ifelse(d$prvdr_num == p, curr_zip, d$zip_code)
}
}

# map zip_codes to states using supplementary data
states <- read.csv("zipcode_state.csv")

states = states[,c(1,4)]

for (zip in 1:length(d$zip_code)) {
d$state[zip] = ifelse(is.na(d$state[zip]) | d$state[zip] == '', states[match(d$zip_code[zip], states$Zipcode),2], d$state[zip])
}

# reformat provider chain identities
d$chain_identity = sub("^DA.{3,5}A.*$", "DAVITA", d$chain_identity)
d$chain_identity = sub("^(FR|FE).{3,9}?S.*$", "FRESENIUS", d$chain_identity)
d$chain_id = ifelse(d$chain_indicator == 'N', 0, ifelse(d$chain_identity == 'DAVITA', 2, ifelse(d$chain_identity == 'FRESENIUS', 3, 1)))

# remake chain_indicator based on chain_identity
d$chain_indicator = ifelse(d$chain_identity == 'DAVITA' | d$chain_identity == 'FRESENIUS', 'Y', d$chain_indicator)
```

# Analysis
Our goal of this analysis was to compare clinics operated by chains with 
independent clinics to find any discrepancies in costs and behaviors. 
To accomplish this, we investigated three questions using the data cleaned 
above.

Reference/Background: "How Acquisitions Affect Firm Behavior and Performance: Evidence From the Dialysis Industry" by Eliason, Paul J.; Heebsh, Benjamin; McDevitt, Ryan C.; Roberts, James W.

### Analysis on EPO Net Cost, Cost, and Rebates
```{r analysis-1}
attach(d)

print("Linear Regression on Indepdents")
avgEpoNetIndependent = glm(epo_net_cost ~ non_medicare_sessions + non_medicare_sessions_indirect + avg_days_open_per_week + avg_session_time + num_machines_regular + dialyser_reuse_times + epo_total + epo_cost + epo_rebates + total_treatments_hd + total_costs_hd_drugs + total_costs_hd_housekeeping + total_costs_hd_machines + total_costs_hd_salaries + total_costs_hd_benefits, data = d, subset = which(d$chain_id == 0))
summary(avgEpoNetIndependent)

print("Linear Regression on Chains")
avgEpoNetChain = glm(epo_net_cost ~ non_medicare_sessions + non_medicare_sessions_indirect + avg_days_open_per_week + avg_session_time + num_machines_regular + dialyser_reuse_times + epo_total + epo_cost + epo_rebates + total_treatments_hd + total_costs_hd_drugs + total_costs_hd_housekeeping + total_costs_hd_machines + total_costs_hd_salaries + total_costs_hd_benefits, data = d, subset = which(d$chain_id != 0))
summary(avgEpoNetChain)

chains = subset(d, d$chain_id != 0)
independent = subset(d, d$chain_id == 0)

par(mfrow=c(2,2))
plot(chains$epo_net_cost, ylim = c(0,5000000), main ="Epo Net Cost for Chains", col = factor(chains$chain_id), pch=19, ylab="Epo Net Cost", xlab = toString(mean(na.omit(chains$epo_net_cost))))
legend("topleft", legend = levels(factor(chains$chain_id)), col = factor(levels(factor(chains$chain_id))), pch=19)
plot(independent$epo_net_cost, ylim = c(0,5000000), main ="Epo Net Cost for Independent", col = factor(independent$chain_id), pch=18, ylab="Epo Net Cost", xlab = toString(mean(na.omit(independent$epo_net_cost))))
boxplot(chains$epo_net_cost~ chains$chain_id, data = chains, ylab = "Epo Net Cost", xlab="chain id", main ="Epo Net Cost for Chains", ylim = c(0,5000000))
boxplot(independent$epo_net_cost ~ independent$chain_id, data = chains, ylab = "Epo Net Cost", xlab="chain id", main ="Epo Net Cost for Independent", ylim = c(0,5000000))


par(mfrow=c(2,2))
boxplot(chains$epo_rebates ~ chains$chain_id, data = chains, ylab = "Epo Rebate", xlab="chain id", main ="Epo Rebate for Chains", ylim = c(0,5000000))
boxplot(independent$epo_rebates ~ independent$chain_id, data = chains, ylab = "Epo Rebate", xlab="chain id", main ="Epo Rebate for Independent", ylim = c(0,5000000))
boxplot(chains$epo_cost~ chains$chain_id, data = chains, ylab = "Epo Cost", xlab="chain id", main ="Epo Cost for Chains", ylim = c(0,5000000))
boxplot(independent$epo_cost ~ independent$chain_id, data = chains, ylab = "Epo Cost", xlab="chain id", main ="Epo Cost for Independent", ylim = c(0,5000000))
```
<br />
We first set out to run separate linear regressions based on chain and 
independent firms and regress them for epo_net_cost against all numeric features.
We found that EPO cost and EPO rebates are strong significant predictors of EPO 
net cost based on our regression. This should come as no surprise, as EPO total 
cost - rebate should give us net cost. From this, we created box plots that 
visualized the EPO rebates and EPO costs for the different chain categories 
and the independent clinics. This showed that, on average, EPO rebates and 
costs were higher for chain clinics than independent ones. However, we also 
found that, on average, the net costs for chains were lower than independent 
counterparts by about $34,000. This aligns with previous studies on EPO, which 
found that dosage more than doubled post-acquisition. 
<br/>
Since Medicare reimbursement for drugs like EPO are determined by quantity 
administered, firms would be incentivized to increase dosage in their patients,
despite medical evidence that excessive EPO increases risk for cardiovascular 
events. Although we cannot conclusively attribute a causal effect to our 
findings, our regression does match the research paper's study. Given more 
granular patient and treatment data, we could better analyze this relationship.

### Analysis on Salaries

```{r analysis-2}
averageOfEachYearIndependent = rep(0, 12)
averageOfEachYearChain = rep(0, 12)
averageOfEachOther = rep(0, 12)
averageOfEachDavita = rep(0, 12)
averageOfEachFresenius = rep(0, 12)
count = 1
for(i in 1998:2010) {
  yearSubsetIndependent = subset(independent, year == i)
  yearSubsetChain = subset(chains, year == i)
  yearSubsetOther = subset(chains, year == i & chain_id == 1)
  yearSubsetDavita = subset(chains, year == i & chain_id == 2)
  yearSubsetFresenius = subset(chains, year == i & chain_id == 3)
  averageOfEachYearIndependent[count] = mean(na.omit(yearSubsetIndependent$total_costs_hd_salaries))
  averageOfEachYearChain[count] = mean(na.omit(yearSubsetChain$total_costs_hd_salaries))
  averageOfEachOther[count] = mean(na.omit(yearSubsetOther$total_costs_hd_salaries))
  averageOfEachDavita[count] = mean(na.omit(yearSubsetDavita$total_costs_hd_salaries))
  averageOfEachFresenius[count] = mean(na.omit(yearSubsetFresenius$total_costs_hd_salaries))
  count = count + 1
}

par(mfrow=c(1,2))
plot(1998:2010, averageOfEachYearChain, pch = 16, col = "red", ylim = c(400000, 700000), ylab ="Average Salary", main ="Independent vs Chain", xlab ="Year") 
par(new = TRUE)                           
plot(1998:2010, averageOfEachYearIndependent, pch = 17, col = "green",axes = FALSE, xlab = "", ylab = "", ylim = c(400000, 700000)) 
legend("topleft",legend = c("chain", "independent"), col=c("red", "green"), pch=c(16,17))

plot(1998:2010, averageOfEachOther, pch = 16, col = "red", ylim = c(400000, 700000), ylab ="Average Salary", main = "By Chain Type", xlab="Year")
par(new = TRUE)
plot(1998:2010, averageOfEachDavita, pch = 17, col = "green", axes = FALSE, xlab = "", ylab = "", ylim = c(400000, 700000))
par(new = TRUE)
plot(1998:2010, averageOfEachFresenius, pch = 18, col = "blue", axes = FALSE, xlab = "", ylab = "", ylim = c(400000, 700000))
par(new = TRUE)
plot(1998:2010, averageOfEachYearChain, pch = 19, col = "black", axes = FALSE, xlab = "", ylab = "", ylim = c(400000, 700000)) 
legend("topleft",legend = c("Other", "Davita", "Fresenius", "Average of all"), col=c("red", "green", "blue", "black"), pch=c(16:19))
```
<br />
We found that there was a consistent difference between the average salaries of
independent clinics compared to chain clinics from 1998 to 2010. Our first graph
illustrates this difference, with independent clinics paying higher salaries on
average than chain clinics on average each year, even with increases in salaries
across the board.
<br />
Our second graph shows how the chain clinics differ in their average salaries. 
The gap is further emphasized, with Davita and Fresenius consistently paying 
at or below the average calculated salary compared to other chains. Overall, 
this aligns with the findings reported in the research paper and podcast,
as researchers concluded that chain clinics were cutting costs by moving towards
more low-skill technicians and less highly-trained nurses. Typically, nurses 
are more qualified and provide better care for patients than technicians, 
which translates into higher salaries. From aggregate salary-collecting websites
like Glassdoor (founded 2007) and Ziprecruiter (founded 2010), dialysis nurse 
salaries in the US have been around \$60-75k over the past decade, while 
dialysis technician salaries have ranged from \$40-50k. These external 
statistics, in tandem with our findings, corroborate with previous research.


### Non-Medicare Sessions

```{r analysis-3}
print("Linear Regression on Non Medicare Sessions for Chains")
summary(glm(non_medicare_sessions ~ total_costs_hd_drugs + total_treatments_hd +
            total_costs_hd_labs + dialyser_reuse_times + total_costs_hd_benefits 
            + total_costs_hd_housekeeping + total_costs_hd_machines 
            + total_costs_hd_other + total_costs_hd_salaries + 
              total_costs_hd_supplies, data = chains))

print("Linear Regression on Non Medicare Sessions for Independents")
summary(glm(non_medicare_sessions ~ total_costs_hd_drugs + total_treatments_hd 
            + total_costs_hd_labs + dialyser_reuse_times + total_costs_hd_benefits
            + total_costs_hd_housekeeping + total_costs_hd_machines +
              total_costs_hd_other + total_costs_hd_salaries + 
              total_costs_hd_supplies, data = independent))

```
We used multiple costs variables to look at changes in the number of sessions 
done through private (non-Medicare) insurance. From our regressions, we can see
that the total cost for hemodialysis drugs has a positive relationship with 
nonmedicare session for chain clinics (coefficient = 2.106e-03), while this 
relationship became negative for independent clinics (coefficient = -1.871e-03).
The same trend can be found for the total 'other' and salary costs for 
hemodialysis, changing from 1.235e-03 to -5.327e-04 and 5.265e-03 to -1.112e-03, 
respectively between chain and independent clinics.
<br />
In general, large dialysis chains have more market and bargaining power, 
and are thus able to negotiate higher reimbursement rates from private insurers.
Chain firms would then be incentivized move towards non-medicare sessions as 
their costs increase, in order to receive more reimbursements. In contrast, 
independent chains are forced to revert to Medicare's flat rates, having less 
bargaining power. Prior studies and articles touch on the role that 
Medicare plays in the dialysis industry, and our findings seem to align with 
these general economic trends.

